Hydraulically-vulnerable trees survive on deep-water access during droughts in a tropical forest

### 2021-05-06 10.0_Supporting_Information.Rmd
Incorporated Production Managers comments

---
title: "Hydraulically-vulnerable trees survive on deep-water access during droughts in a tropical forest"
author: ''
date: ''
output:
  bookdown::word_document2:
        reference_docx: manuscriptstyle.docx
        fig_caption: yes
        number_sections: false
  bookdown::html_document2:
    number_sections: no
    theme: flatly
    fig_caption: yes
  bookdown::pdf_document2: default
bibliography: references.bib
csl: new-phytologist.csl
---

@@

# Supporting Information

## Supplementary Figures

```{r gpp-vpd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "35%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-gpp-vpd}Scatterplot of mean midday (11:00 am to 15:00 pm) gross primary productivity (GPP) and mean midday vapor pressure deficit (VPD) measured at the eddy flux tower in Barro Colorado Island over 2012-2017. Red line shows the fit from a polynomial model described in inset."}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/gpp_by_env_variables/monthly_gpp.vpd.jpeg"))
```

\newpage
```{r ksat-profiles, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ksat-profiles}Generated 5000 curves of soil hydraulic conductivity variation by depth based on Latin Hypercube Sampling within 95% CI of observed data at depths of 12.5 cm and 60 cm [@Kinner:2004], assuming a linear decline between those depths, and values before 12.5 and after 60 cm as that of 12.5 cm and 60 cm, respectively. Only 100 curves are shown for clarity. Note the log scale on horizontal axis.")}
knitr::include_graphics(file.path("figures/Bootstrapped Ksat profiles within observed 95CI_random_two_hundred.jpeg"))
```

\newpage
```{r iso-data, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-iso-data}Stable isotopic ratio ($\\mathrm{\\delta}^2H$) of soil water by depth measured at Barro Colorado Island in March 1997. Data from @Meinzer:1999. $\\mathrm{\\delta}^2H$ to depth relationship was particularly uncertain for $\\mathrm{\\delta}^2H$ values > -40 ‰. Two species, *Guapira standleyana* (-28.9 $\\pm$ 3.7; mean $\\pm$ SE) and *Spondias radlkoferi* (-33.5 $\\pm$ 2.6), in @Meinzer:1999 dataset had $\\mathrm{\\delta}$^2^H~xylem~ values > -40 ‰. For example, *Guapira standleyana*, with $\\mathrm{\\delta}$^2^H value of -28.9 $\\pm$ 3.7 could have a source-water depth anywhere between 10 to 50 cm. *Spondias radlkoferi* is also typically leafless in March-April. These two species were not included in the ERD model-data comparison."}
knitr::include_graphics(file.path("figures/UDI_confidence/Meinzer_etal_1999_Depth_vs_soil_deltaD.jpeg"))
```

\newpage
```{r sens-sm-et, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-sens-sm-et}Parameter sensitivity of ELM-FATES soil water content (that is, VWC, **a**-**b**) and ET (**c**-**d**) is shown for two representative sample months, as sensitivity for a particular parameter varies depending on the chosen month. Horizontal axes show global ranges for a given parameter. Parameters are color coded and are consistent across panels. Each point represents a simulation among the 5000 ensemble-member runs. See S\\@ref(tab:param) for description, global ranges and chosen ranges for parameters."}
knitr::include_graphics(file.path("figures/ELM_FATES_parameter_sensitivity/sens.sm.et.jpeg"))
```

\newpage
```{r sens-run, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-sens-run}Parameter sensitivity of ELM-FATES stream discharge (QRUNOFF) is shown for four sample months (**a**-**d**), as sensitivity for a particular parameter varies depending on the chosen month. Horizontal axes show global ranges for a given parameter. Parameters are color coded and are consistent across panels. Each point represents a simulation among the 5000 ensemble-member runs. See S\\@ref(tab:param) for description, global ranges and chosen ranges for parameters."}
knitr::include_graphics(file.path("figures/ELM_FATES_parameter_sensitivity/sens.run.jpeg"))
```

\newpage
```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Daily mean soil water potential ($\\Psi_{\\textrm{soil},z}$) by depth $z$ (m; panels) over 1990-2018 predicted from each of the the 100 best-fit ensembles (colored lines). Vertical gray lines demarcate median dates of the demographic censuses at the BCI 50 ha plot."}
knitr::include_graphics(file.path("figures/2019-10-14_5000/best-fits/psi_model_daily_all_depths_params.top.few_full_beg_1991.jpeg"))
```

\newpage
```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}Vulnerability curves fitted to observed $K_\\textrm{leaf}$ vs. $\\mathrm{\\Psi}$~leaf~ data for ", length(sp.exp.param$sp)," tree species from BCI--in absolute terms (**a**, **b**), as well as, in terms of loss of conductivity (%) (**c**, **d**). Each line represents a species, color coded by WSG (**a**, **c**), or LMA (**b**, **d**), with blue tones indicating smaller values of WSG or LMA and red tones indicating larger values.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/plot.obs.plc.jpeg"))
```

\newpage
```{r obs-comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-comm-plc-curves}Percent loss of vulnerability curves for species with ERD estimates. Each line represents a species, color coded by source of the curve: fitted to observed data for $K_\\textrm{leaf}$ vs. $\\mathrm{\\Psi}$~leaf~ are in yellow and those based on the scaling relationship with WSG and LMA are in gray.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/obs.mod_plccurves_fits_ggplot_no_col.jpeg"))
```

\newpage
```{r other-erd-mods, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-other-erd-mods}Effective Rooting Depth versus $\\mathrm{\\delta}^2H$ for each of the ERD models tested (panels). For each ERD model only those species have an estimate of ERD that have passed the test of Pearson's $r$ > 0.71 during ERD parameterisation. The number of species with an ERD estimate as well as $\\mathrm{\\delta}^2H$ for each model are shown in inset, along with the goodness-of-fit ($R$^2^) and $p$-values for a linear relationship between ERD and $\\mathrm{\\delta}^2H$. ERD models in panels **a**-**e** correspond to Eq. \\@ref(eq:M1)-\\@ref(eq:M6), respectively.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_sp_color_Meinzer.jpeg"))
```

\newpage
```{r erd-hyd-full, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd-full}Pairwise Spearman's correlations (upper triangle) and corresponding linear regressions (lower triangle) among modelled Effective Rooting Depth (ERD) and hydraulic properties for seven canopy species found on Barro Colorado Island (Panama).Spearman's correlation coefficients and significance levels are shown (upper-half; '*' $p$ < 0.05; '.' $p$ < 0.1; ' ' $p$ < 1).")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/erd.stem.traits_cor.chart.jpeg"))
```

\newpage
```{r erd-mort-deci, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-cdeci}Mortality rates across census intervals from 1985-2015 for species (circles) of various deciduous leaf-habit categories (colors) plotted against ERD. $R$^2^ and significance levels for linear model fits are given in panel insets. Census-intervals with a superscript * are shown for comparison and indicate intervals in which significant mortality was explained by ERD among evergreen species (see Fig. 8 in the main text)."}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_by rdi.gr_deciduous.jpeg"))
```

\newpage
```{r erd-time-deci, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time-deci}Modeled effective rooting depth (ERD; horizontal-axis) versus time spent beyond critical hydraulic threshold (vertical axis) by census interval (colored bars) for deciduous species in mortality analyses (Fig. \\@ref(fig:erd-mort-deci)). Each bar represents the average time species of the same ERD spent beyond species-specific critical hydraulic thresholds in a given interval, that is, the proportion of days for which $\\mathrm{\\Psi}_{\\textrm{soil},z=\\textrm{ERD}}$ was more negative than species $\\mathrm{\\Psi}$~crit~, defined as $\\mathrm{\\Psi}$~20,leaf~, and where $z$ is the soil depth matching species ERD. Standard error of the mean are shown over each bar. Note the squared y-axis scale.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mfac vs. rdi.gr_deciduous.jpeg"))
```

## Supplementary Tables

```{r param, results = 'asis', echo = FALSE}
knitr::kable(param.table, caption = 'ELM-FATES parameters used to generate ensembles, with description, prescribed global ranges, rationale for the choice of ranges and references. Ranges for best-fit chosen ensembles are also given. Note that no. 1-6 are parameters specific to FATES.', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(param.table))))
```

\newpage
```{r hyd-table, results = 'asis', echo = FALSE}
knitr::kable(hyd.table, caption = paste0("\\label{lab-hyd-table} Minimum leaf water potential $\\mathrm{\\Psi}$~min~, leaf turgor loss point $\\mathrm{\\Psi}$~tlp~, stem maximum hydraulic conductivity $K_\\textrm{max,stem}$ and stem vulnerability to cavitation $\\mathrm{\\Psi}$~88,stem~ for seven species with ERD estimates. $\\mathrm{\\Psi}$~min~ is sourced from @Wolfe.2019, while $\\mathrm{\\Psi}$~tlp~, $K_\\textrm{max,stem}$, $\\mathrm{\\Psi}$~50,stem~ and $\\mathrm{\\Psi}$~88,stem~ are sourced from @Wolfe.2021. $\\mathrm{\\Psi}$~min~, $\\mathrm{\\Psi}$~tlp~, $\\mathrm{\\Psi}$~50,stem~ and $\\mathrm{\\Psi}$~88,stem~ are in MPa, while $K_\\textrm{max,stem}$ in kg m^-1^ s^-1^ MPa^-1^."), booktabs = TRUE, escape = FALSE)
```

\newpage
```{r ab-param, results = 'asis', echo = FALSE}
knitr::kable(ab.table, caption = paste0("\\label{lab-ab-param}Leaf vulnerability curves parameters *A* & *B*, $K_\\textrm{max,leaf}$ and $\\mathrm{\\Psi}$~20,leaf~ for the ", length(erd.sp), " species with ERD estimates. Species for which curves were fitted to observed data for $K_\\textrm{leaf}$ vs. $\\mathrm{\\Psi}$~leaf~ are identified with column Source as Data, while species for which curves were obtained based on trait-proxies are identified as Model."), booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(ab.table))))
```

\newpage
## Additional details of the ELM-FATES Model

In ELM, soil water is predicted from a multi-layer model, in which the vertical soil moisture transport is governed by infiltration, surface and sub-surface runoff, gradient diffusion, gravity, canopy transpiration through root extraction, and interactions with groundwater (see @Oleson.2013 for process details and equations). FATES simulates size-structured tree community complete with individual tree rooting profiles. In our use of reduced-complexity configuration of ELM-FATES with a single plant functional type, ELM removes total community canopy transpiration as water-uptake from the soil layers in proportion to the total root fraction in each layer in the 1-D column. Note that we use the soil water potential dynamics obtained from ELM-FATES as an ecosystem-level outcome to drive our species-level, empirical ERD model, and thus believe that there is no conflict/circularity between tree-level rooting profiles or water uptake processes in ELM-FATES and the ERD model.

\
We ran ELM with FATES vegetation, in which the ELM model simulates interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\mathrm{\Delta}W_{\textrm{can}}$, surface water $\mathrm{\Delta}W_{\textrm{sfc}}$, soil water $\mathrm{\Delta}W_{\textrm{liq},z_{i}}$ at depth $z_{i}$ and water in an unconfined aquifer $\mathrm{\Delta} W_{a}$ (omitting processes relevant to snow, wetlands or lakes). Conservation of these terms is expressed as follows:

\
Justification for the use of static stand structure: In static stand structure model, the model is started from an observed forest inventory and recruitment, growth and mortality are deactivated. Holding the stand structure static rather than allowing it to emerge from physiologically-mediated competition between trees removes a set of internal model feedbacks and thus the only differences between model simulations are driven by direct physiological effects of perturbed model parameters. This allows us to focus on the hydraulic and hydrological questions explored in this paper.

## Additional details for ELM-FATES model parameterization

In the Conrad catchment, in which the 50 ha plot is situated, soil hydraulic conductivity decreases with depth from 12.5 and 60 cm (@Godsey:2004; Table \@ref(tab:param)). We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths <12.5 cm and >60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig. \@ref(fig:ksat-profiles)).   
\
For greater accuracy, instead of the ELM default pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content (GWC) vs. $\mathrm{\Psi}$~soil~ for depths of 0.15, 0.4 and 1 m [@Kupers:2019]. We converted GWC to volumetric (VWC, $\theta$) using a measured bulk density value of 0.8 g cm^-3^ (see below). The @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences in water retention among depths were nominal,

## ELM-FATES calibration

We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq. \@ref(eq:water-balance), namely, (1) evapotranspiration ET from the flux tower by the 50 ha plot ($\backsimeq$ $E_{v} + E_{g}$ in Eq. \@ref(eq:water-balance); 2012-2017; see [SI](#SIData.matteo))), (2) local stream discharge $q_\textrm{run}$ ($\backsimeq$ $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see @Paton.2019 and [SI](#SIData.steveP)) and (3) soil volumetric water content (VWC) $\mathrm{\theta}_{z}$ ($\backsimeq$ $W_{\textrm{liq,} z}$) from two sources: (i) a long-term (2012-2018) record of VWC $\mathrm{\theta}_\textrm{0-0.15 m}$ averaged across three vertical time domain reflectometry (TDR) probes over depth 0-15 cm from three locations near the flux tower, and (ii) plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 at depths of 0.15, 0.4 and 1 m (1299 samples covering all soil types and habitats; see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter, $\mathrm{\theta}_{\textrm{0-0.15 m}}$, $\mathrm{\theta}_{\textrm{0.4 m}}$ and $\mathrm{\theta}_{\textrm{1 m}}$ refer to plot-wide averages).

\
For ELM-FATES calibration we calculated an objective function for each of the 5000-member ensemble by equally weighting standardized RMSE between observations and simulations across all fluxes and states mentioned above. We describe the objective function $O$ as:

\
Model fit to soil moisture dynamics by depth was one of our key priority. Preliminary analyses showed that fitting the model to all states and fluxes simultaneously captured soil moisture dynamics for shallower depths but not at 1 m depth. To ensure model fit at 1 m depth, we thus imposed a two-step procedure, wherein we first chose the 300 parameter ensemble members that minimized $\textrm{RMSE}^{*}({\mathrm{\theta}_{\textrm{1 m}}}, W_{\textrm{liq,`r soil.depths[8]` m}})$ (< 0.2), and among those 300 ensemble members we chose 100 with the lowest values for the objective function $\overline{O}$.

## Alternative model structures for Effective Rooting Depth (ERD)

>>>>

---
title: ""
author: ''
date: ''
output:
  bookdown::word_document2:
        reference_docx: manuscriptstyle_without_line_number.docx
        fig_caption: false
        number_sections: false
  bookdown::html_document2:
    number_sections: no
    theme: flatly
    fig_caption: yes
  bookdown::pdf_document2: default
bibliography: references.bib
csl: new-phytologist.csl
indent: true
header-includes:
    - \usepackage{caption}
    - \usepackage{indentfirst}
    - \setlength{\parindent}{24pt}
    # - \setlength{\parskip}{0em}
---

\captionsetup[table]{labelformat=empty}
\usepackage{float}

@@

_**New Phytologist**_ Supporting Information

**Hydraulically-vulnerable trees survive on deep-water access during droughts in a tropical forest**

Chitra-Tarak, Rutuja, Chonggang Xu, Salomón Aguilar, Kristina J. Anderson-Teixeira, Jeff Chambers, Matteo Detto, Boris Faybishenko, Rosie A. Fisher, Ryan G. Knox, Charles D. Koven, Lara M. Kueppers, Nobert Kunert, Stefan J. Kupers, Nathan G. McDowell, Brent D. Newman, Steven R. Paton, Rolando Pérez, Laurent Ruiz, Lawren Sack, Jeffrey M. Warren, Brett T. Wolfe, Cynthia Wright, S. Joseph Wright, Joseph Zailaa, Sean M. McMahon

\
Article acceptance date: 29 April 2021

\
The following Supporting Information is available for this article:

**Figure S1**  Observed relationship between GPP and VPD at Barro Colorado Island.  
**Figure S2**  Seasonal variation in estimated standardised LAI for 29 species at Barro Colorado Island.  
**Figure S3**  Ensembles of soil hydraulic conductivity variation by depth for Conrad Trail Stream Catchment, Barro Colorado Island. Data from @Godsey:2004.  
**Figure S4**  Volumetric water content vs. TDR data for a vertical TDR probe used for TDR probe calibration.  
**Figure S5**  $\mathrm{\delta}$^2^H of soil water by depth measured at Barro Colorado Island. Data from @Meinzer:1999.   
**Figure S6**  Parameter sensitivity of ELM-FATES soil water content and evapo-transpiration.  
**Figure S7**  Parameter sensitivity of ELM-FATES stream discharge.  
**Figure S8**  ELM-FATES predicted soil water potential dynamics by depth over 1990-2018.  
**Figure S9**  Leaf hydraulic vulnerability curves fitted to observed $K_\textrm{leaf}$ vs. $\mathrm{\Psi}$~leaf~ data for 21 tree species from Barro Colorado Island.  
**Figure S10**  Percent loss of leaf hydraulic conductivity curves for ERD species, including curves fitted to $K_\textrm{leaf}$ vs. $\mathrm{\Psi}$~leaf~ as well as those based on scaling relationships with leaf mass area and wood specific gravity.  
**Figure S11**  Effective rooting depth (ERD) versus $\mathrm{\delta}$^2^H~xylem~ for each of the ERD model structures tested.   
**Figure S12**  Correlation matrix between ERD and hydraulic traits.  
**Figure S13**  Mortality rates for deciduous species across census intervals from 1985-2015 versus ERD.  
**Figure S14**  Modeled effective rooting depths of deciduous species versus time spent beyond critical hydraulic threshold.

**Table S1**  ELM-FATES parameters used to generate ensembles, with description, prescribed global ranges, rationale for the choice of ranges and references, as well as ranges for best-fit ensembles.  
**Table S2**  QA/QC procedure applied to eddy covariance fluxes.  
**Table S3**  Above-ground hydraulic traits data from @Wolfe.2019 and @Wolfe.2021 used for comparison with ERD.   
**Table S4**  Leaf vulnerability curve parameters A & B, $K_\textrm{max,leaf}$ and $\mathrm{\Psi}_\textrm{20,leaf}$ by source (data or model) for the 29 species with ERD estimates.

**Methods S1** Alternative structures for effective rooting depth model.  
**Methods S2** Statistics for identifying best-fit ERD.   
**Methods S3** Processing of forest census data for growth estimates.     
**Methods S4** Leaf Area Index calculations.   
**Methods S5** Details of the ELM-FATES model.  
**Methods S6** Details for ELM-FATES model parameterization.  
**Methods S7** ELM-FATES calibration.

**Dataset S1** Microclimatic and flux tower data.   
**Dataset S2** Leaf hydraulic conductivity and vulnerability to cavitation.   
**Dataset S3** Wood specific gravity (WSG) and leaf mass area (LMA).  
**Dataset S4** Leaf deciduousness categories.  
**Dataset S5** Stream discharge from the Conrad catchment.  
**Dataset S6** Volumetric water content.  
**Dataset S7** Stem maximum hydraulic conductivity and vulnerability to cavitation.  
**Dataset S8** Leaf turgor loss point.  
**Dataset S9** Above-ground hydraulic safety margins.

**Notes S1** ERD model structure selection.  
**Notes S2** Additional results for exposure to water-stress.

\newpage
**Figure S1** Scatterplot of mean midday (11:00 am to 15:00 pm) gross primary productivity (GPP) and mean midday vapor pressure deficit (VPD) measured at the eddy flux tower in Barro Colorado Island over 2012-2017. Red line shows the fit from a polynomial model described in inset.

```{r gpp-vpd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "35%", fig.width = 3, fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-gpp-vpd}"}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/gpp_by_env_variables/monthly_gpp.vpd.jpeg"))
```

\newpage
**Figure S2** Seasonal variation in standardised LAI (range 0-1) for `r length(erd.sp)` species at Barro Colorado Island (panels). Species are color-coded by leaf habit as defined by expert botanists.

```{r leaf-cover, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-leaf-cover}")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf.cover_BCI_multi_panel.jpeg"))
```

\newpage
**Figure S3** Generated 5000 curves of soil hydraulic conductivity variation by depth based on Latin Hypercube Sampling within 95% CI of observed data at depths of 12.5 cm and 60 cm for Conrad Trail Stream Catchment [@Godsey:2004], assuming a linear decline between those depths, and values before 12.5 and after 60 cm as that of 12.5 cm and 60 cm, respectively. Only 100 curves are shown for clarity. Note the log scale on horizontal axis.

```{r ksat-profiles, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ksat-profiles}")}
knitr::include_graphics(file.path("figures/Bootstrapped Ksat profiles within observed 95CI_random_two_hundred.jpeg"))
```

\newpage
**Figure S4** Volumetric water content vs. TDR data for a vertical TDR probe (0-15 cm depth) at the base of the eddy flux tower near the 50 ha plot at BCI.

```{r TDR-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-TDR-calib}"}
knitr::include_graphics(file.path("figures/Matteos_TDR_SWC_data_linear_model.jpeg"))
```

\newpage
**Figure S5** Stable isotopic ratio ($\mathrm{\delta}^2H$) of soil water by depth measured at Barro Colorado Island in March 1997. Groundwater sampled at the sources of springs on BCI during this period yielded an average $\mathrm{\delta}^2H$ value of –60 ‰ (not shown). Data from @Meinzer:1999. $\mathrm{\delta}^2H$ to depth relationship was particularly uncertain for $\mathrm{\delta}^2H$ values > -40 ‰. Two species, *Guapira standleyana* (-28.9 $\pm$ 3.7; mean $\pm$ SE) and *Spondias radlkoferi* (-33.5 $\pm$ 2.6), in @Meinzer:1999 dataset had $\mathrm{\delta}$^2^H~xylem~ values > -40 ‰. For example, *Guapira standleyana*, with $\mathrm{\delta}$^2^H~xylem~ value of -28.9 $\pm$ 3.7 could have a source-water depth anywhere between 10 to 50 cm. *Spondias radlkoferi* is also typically leafless in March-April. These two species were not included in the ERD model-data comparison.

```{r iso-data, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-iso-data}"}
knitr::include_graphics(file.path("figures/UDI_confidence/Meinzer_etal_1999_Depth_vs_soil_deltaD.jpeg"))
```

\newpage
**Figure S6** Parameter sensitivity of ELM-FATES soil water content (that is, VWC, **a**-**b**) and evapotranspiration (ET, **c**-**d**) is shown for two representative sample months, as sensitivity for a particular parameter varies depending on the chosen month. Horizontal axes show global ranges for a given parameter. Parameters are color coded and are consistent across panels. Each point represents a simulation among the 5000 ensemble-member runs. See Table S\@ref(tab:param) for description, global ranges and chosen ranges for parameters.

```{r sens-sm-et, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-sens-sm-et}"}
knitr::include_graphics(file.path("figures/ELM_FATES_parameter_sensitivity/sens.sm.et.jpeg"))
```

\newpage
**Figure S7** Parameter sensitivity of ELM-FATES stream discharge (QRUNOFF) is shown for four sample months (**a**-**d**), as sensitivity for a particular parameter varies depending on the chosen month. Horizontal axes show global ranges for a given parameter. Parameters are color coded and are consistent across panels. Each point represents a simulation among the 5000 ensemble-member runs. See Table S\@ref(tab:param) for description, global ranges and chosen ranges for parameters.

```{r sens-run, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-sens-run}"}
knitr::include_graphics(file.path("figures/ELM_FATES_parameter_sensitivity/sens.run.jpeg"))
```

\newpage
**Figure S8** Daily mean soil water potential ($\Psi_{\textrm{soil},z}$) by depth $z$ (in m; panels) over 1990-2018 predicted from each of the 100 best-fit ensembles (colored lines). Vertical gray lines demarcate median dates of the tree demographic censuses at the BCI 50 ha plot.

```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}"}
knitr::include_graphics(file.path("figures/2019-10-14_5000/best-fits/psi_model_daily_all_depths_params.top.few_full_beg_1991.jpeg"))
```

\newpage
**Figure S9** Leaf hydraulic vulnerability curves fitted to observed $K_\textrm{leaf}$ vs. $\mathrm{\Psi}$~leaf~ data for `r length(sp.exp.param$sp)` tree species from BCI--in absolute terms (**a**, **b**), as well as, in terms of percent loss of conductivity (**c**, **d**). Each line represents a species, color coded by wood specific gravity (WSG) (**a**, **c**), or leaf mass area (LMA) (**b**, **d**), with blue tones indicating smaller values of WSG or LMA and red tones indicating larger values.

```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/plot.obs.plc.jpeg"))
```

\newpage
**Figure S10** Percent loss of leaf hydraulic conductivity curves for species with ERD estimates. Each line represents a species, color coded by source of the curve: fitted to observed data for $K_\textrm{leaf}$ vs. $\mathrm{\Psi}$~leaf~ are in yellow (data) and those based on the scaling relationships with leaf mass area and wood specific gravity are in gray (model).

```{r obs-comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-comm-plc-curves}")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/obs.mod_plccurves_fits_ggplot_no_col.jpeg"))
```

\newpage
**Figure S11** Effective rooting depth versus $\mathrm{\delta}^2H$~xylem~ for each of the ERD model structures tested (panels). For each ERD model structure only those species have an estimate of ERD that have passed the test of Pearson's $r$ > 0.71 during ERD parameterisation. The number of species with an ERD estimate as well as $\mathrm{\delta}^2H$ for each model structure are shown in inset, along with the goodness-of-fit ($R$^2^) and $p$-values for a linear relationship between ERD and $\mathrm{\delta}^2H$~xylem~. ERD models in panels **a**-**e** correspond to Eq. S\@ref(eq:M1)-S\@ref(eq:M6), respectively.

```{r other-erd-mods, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-other-erd-mods}")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_sp_color_Meinzer.jpeg"))
```

\newpage
**Figure S12** Pairwise Spearman's correlations (upper triangle) and corresponding linear regressions (lower triangle) among modelled effective rooting depth (ERD) and hydraulic properties for seven canopy species found on Barro Colorado Island (Panama). Spearman's correlation coefficients and significance levels are shown (upper triangle; '$**$' $p$ < 0.01, '$*$' $p$ < 0.05, '$.$' $p$ < 0.1).

```{r erd-hyd-full, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd-full}")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/erd.stem.traits_cor.chart.jpeg"))
```

\newpage
**Figure S13** Mortality rates across census intervals from 1985-2015 for species (circles) of various deciduous leaf-habit categories (colors) plotted against ERD. $R$^2^ and significance levels for linear model fits are given in panel insets. Census-intervals with a superscript $*$ and $**$ are shown for comparison with a corresponding figure for evergreen species (Fig. 8 in the main text) and indicate intervals in which significant mortality was explained by ERD among evergreen species at $\alpha$ = 0.1 and $\alpha$ = 0.05, respectively.

```{r erd-mort-deci, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-cdeci}"}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_by rdi.gr_deciduous.jpeg"))
```

\newpage
**Figure S14** Modeled effective rooting depth (ERD; horizontal-axis) versus time spent beyond critical hydraulic threshold (vertical axis) by census interval (colored bars) for deciduous species in the mortality analyses (Fig. S\@ref(fig:erd-mort-deci)). Each bar represents the average time species of the same ERD spent beyond species-specific critical hydraulic thresholds in a given interval, that is, the proportion of days for which $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ was more negative than species $\mathrm{\Psi}$~crit~, defined as $\mathrm{\Psi}$~20,leaf~, and where $z$ is the soil depth matching species ERD. Standard error of the mean are shown over each bar. Note the squared y-axis scale.

```{r erd-time-deci, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time-deci}")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mfac vs. rdi.gr_deciduous.jpeg"))
```

\newpage
**Table S1** ELM-FATES parameters used to generate ensembles, with parameter description, prescribed global ranges, rationale for the choice of ranges and references. Ranges for best-fit chosen ensembles are also given. Note that the first six parameters are specific to FATES.

```{r param, results = 'asis', echo = FALSE, tab.cap = NULL}
knitr::kable(param.table, caption = "", booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(param.table))))
```

**Table S2** QA/QC procedure applied to eddy covariance fluxes. w: vertical wind speed; t: air temperature; q: water vapor density; c: CO~2~ density; u: horizontal wind speed. The bar indicates 30-min averaging and the carat indicates fluctuation ($\overline{\hat{x} = x - \bar{x}}$)

```{r eddy, results = 'asis', echo = FALSE}
knitr::kable(eddy.table, caption = "", booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(eddy.table))))
```
